---
title: "NBA Daily Fantasy Exploration"
author: "Ian Whitestone"
date: "October 16, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Daily fantasy sports (DFS) are a subset of fantasy sports, where participants construct lineups based on the games occuring on a given day [<https://en.wikipedia.org/wiki/Daily_fantasy_sports>]. Lineups are subject to various constraints, such as a salary cap and having a minimum number of players at each position. As with other fantasy sports, player's fantasy points are based on their actual performance in real life. As a result, a key component of being a successful DFS player is the ability to project a player's total points for a given night. 

The study focuses on the NBA DFS, with a particular emphasis on examining factors that influence a player's score. The NBA data used in the study is sourced from Erik Berg's API  [<https://erikberg.com/api>]. The study covers NBA data from the 2012-2013 season, up until the 2015-2016 season. Fantasy points are calculated using Fanduel's scoring system, described below. Unlike other sports, all positions in the NBA are scored using the same system.

* 3-pt FG = 3pts
* 2-pt FG = 2pts
* FT = 1pt
* Rebound = 1.2pts
* Assist = 1.5pts
* Block = 2pts
* Steal = 2pts
* Turnover = -1pt

Source: <https://www.fanduel.com/rules>

##

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
http://vidzi.tv/j4me09rfrxg8.html